Analysis-ready VCF at Biobank-scale using Zarr

### 2024-05-07 paper.tex
More updates

VCF is a text-based format in which each line
describes observations
and metadata for a single variant~\citep{danecek2011variant}. 
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF (a packed binary 
representation) is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage formats such as Parquet~\citep{parquet2024},
makes retrieving particular columns much 
more efficient and can lead to substantially better compression.
While columnar techniques have been successfully applied 
alignment
% TODO check
% Jeff: OK with the ADAM reference in here? You did some parquet based
% alignment storage, right?
storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
use of technologies such as Parquet for 
storing and analysing variation data have had limited
success~\citep{boufea2017managing,fan2020variant}.
Directly mapping VCF data to a columnar layout, in which there is a 
a column for the genotypes (and other per-sample QC metrics) 
for each sample, scales poorly. Firstly, it leads to a large number 
of columns (millions, at Biobank scale), which techologies such 
as Parquet are not designed for. Secondly, compression performance 
within columns is poor because there is much less similarity among
the genotypes for a given sample than there is among the genotypes
for a given variant. [TODO check this - reference?]

@@

>>>>

VCF is row-wise format in which 
describes observations
and metadata for a single variant~\citep{danecek2011variant} are
encoded as a line of text.
BCF (a packed binary representation) is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).
The row-wise approach works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.

The usual alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage formats such as Parquet~\citep{parquet2024},
make retrieving particular columns much 
more efficient and can lead to substantially better compression.
While columnar techniques have been successfully applied 
in alignment
% TODO check
% Jeff: OK with the ADAM reference in here? You did some parquet based
% alignment storage, right?
storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
the use of technologies such as Parquet for 
storing and analysing variation data have had limited
success~\citep{boufea2017managing,fan2020variant}.
Columnar techologies tend to be 
optimised for ``tall and narrow'' data there is a large number of records
for a relatively small number of columns. 
Mapping VCF directly to a columnar layout, in which there is a 
a column for the genotypes (and other per-sample QC metrics) 
leads to millions of columns, and substantial technical challenges.
Additionally, compression performance within columns is poor 
because there is much less similarity among
the genotypes for a given sample than there is among the genotypes
for a given variant. [TODO check this - reference?]

@@

VCF is at its core an encoding of the genotype matrix $G$, where each entry
encodes the observed genotypes for a given sample at a give variant site,
interleaved with per-variant information
and other matrices (e.g., GQ or AD).
The data is largely numerical and of a fixed dimension, and is therefore a 
natural mapping to an array-oriented approach [PICK UP HERE]
